---
title: "Interactive Shiny report of HTSanalyzeR2"
author: 
- name: Lina Zhu 
  affiliation: Department of Biomedical Sciences, City University of Hong Kong, Hong Kong
- name: Feng Gao
  affiliation: Department of Biomedical Sciences, City University of Hong Kong, Hong Kong
- name: Xiupei Mei
  affiliation: Department of Computer Science, City University of Hong Kong, Hong Kong
- name: Xin Wang
  affiliation: Department of Biomedical Sciences, City University of Hong Kong, Hong Kong 
date: "`r Sys.Date()`"
package: HTSanalyzeR2
output: 
        BiocStyle::pdf_document
bibliography: bibliography1.bib

vignette: >
  %\VignetteIndexEntry{Shiny report of HTSanalyzeR2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this short tutorial, we give a detailed illustration for the shiny report part of HTSanalyzeR2 package to visualize the results and modify figures in different aspects.

# Interactive Shiny report visualization

## Visualize single *GSCA* object for single data set

For single data set analysis, after analyzed by HTSanalyzeR2, we can get a gsca object and use the function *report* to launch the shiny report. 
```{r, eval=FALSE}
data(d7_gsca)
report(d7_gsca)
```

## Visualize single *NWA* object for single data set
```{r}
data(d7_nwa)
report(d7_nwa)
```


## Visualize a list of *GSCA* object for time series/comparative analysis
```{r}
data(gscaTS)
reportAll(gscaTS)
```

## Visualize a list of *NWA* object for time series/comparative analysis
```{r}
data(nwaTS)
reportAll(nwaTS)
```






